Fitness effects of mutations to SARS-CoV-2 proteins

### 2023-01-31 paper/paper.tex

\begin{abstract}
Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and understanding of the functions of viral proteins.
However, experimentally measuring effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two SARS-CoV-2 proteins.
Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate effects of mutations.
We first calculate how many independent occurrences of each mutation are expected to be observed along the SARS-CoV-2 phylogeny in the absence of selection.
We then compare these expected observations to the actual observations to estimate the effect of each mutation.
These estimates correlate well with deep mutational scanning measurements.
For most genes, synonymous mutations are nearly neutral, stop-codon mutations are deleterious, and amino-acid mutations have a range of effects.
However, some viral accessory proteins are under little to no selection.
To enable our estimates to be easily used, we provide interactive visualizations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that many independent occurrences of each neutral mutation are observed.
\end{abstract}

@@

\section{Acknowledgments}
We thank Angie Hinrichs for assistance with questions related to use of the UShER package and its pre-built mutation-annotated tree.
The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
JDB is an Investigator of the Howard Hughes Medical Institute.

>>>>

\begin{abstract}
Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and understanding of the functions of viral proteins.
However, experimentally measuring effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two SARS-CoV-2 proteins.
Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate effects of mutations.
We first calculate how many independent occurrences of each mutation are expected to be observed along the SARS-CoV-2 phylogeny in the absence of selection.
We then compare these expected observations to the actual observations to estimate the effect of each mutation.
These estimates correlate well with deep mutational scanning measurements.
For most genes, synonymous mutations are nearly neutral, stop-codon mutations are deleterious, and amino-acid mutations have a range of effects.
However, some viral accessory proteins are under little to no selection.
We provide interactive visualizations of effects of mutations to all SARS-CoV-2 proteins (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that many independent occurrences of each neutral mutation are observed.
\end{abstract}

@@

\section{Acknowledgments}
We thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built mutation-annotated tree.
The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
JDB is an Investigator of the Howard Hughes Medical Institute.
